The goal of this script is to perform Delong AUC test for bird-location vs. observer-location Boosted Regression Tree prediction results
We want to test if bird-location AUC performance is higher than observer-location

```{r}
# Load necessary library
library(pROC)
library(readr)
library(gbm)
library(dplyr)
set.seed(123)

```

***Species-centered Test Data***
Load Test Data for species-centered
```{r}
# Define the base directory
testdata_dir <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TestData/SpeciesCentered/"

# Define file names
file_names <- c("SpeciesCentered_FixedRadius.csv", "SpeciesCentered_noRadius.csv", "SpeciesCentered_ESW.csv")

# Load and assign datasets to variables
fixed.test_bird <- read.csv(paste0(testdata_dir, file_names[1]))
noRadius.test_bird <- read.csv(paste0(testdata_dir, file_names[2]))
esw.test_bird <- read.csv(paste0(testdata_dir, file_names[3]))


```

Load the Trained models for species-centered strategy
```{r}

model.dir.noradius_birdcentered <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SpeciesCentered/SpeciesCentered_NoRadius"
noradius.models_bird <- list.files(model.dir.noradius_birdcentered, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.fixedradius_birdcentered <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SpeciesCentered/SpeciesCentered_FixedRadius"
fixedradius.models_bird <- list.files(model.dir.fixedradius_birdcentered, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.esw_birdcentered <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SpeciesCentered/SpeciesCentered_ESW"
esw.models.models_bird <- list.files(model.dir.esw_birdcentered, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

```

***Observer-centered Test Data***
Read ESW test dataset
# We will treat esw file differently since parsing issue has occurred
```{r}

esw.test <- readr::read_csv("/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TestData/SurveyCentered/SurveyCentered_ESW.csv", col_names = TRUE)
spec(esw.test)
problems(esw.test)

esw.test_survey <- esw.test[-c(39007,39008),]
```

Load Test Data for observer-centered
```{r}
# Define the base directory
testdata_dir_survey <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TestData/SurveyCentered/"

# Define file names
file_names_survey <- c("SurveyCentered_FixedRadius.csv", "SurveyCentered_noRadius.csv")

# Load and assign datasets to variables
fixed.test_survey <- read.csv(paste0(testdata_dir_survey , file_names_survey[1]))
noRadius.test_survey <- read.csv(paste0(testdata_dir_survey , file_names_survey[2]))

```

Load the Trained models for observer-centered strategy
```{r}
individual_species_oregon <- split(fixed.test_survey, fixed.test_survey$Common_Name)

model.dir.noradius_survey <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_NoRadius"
noradius.models_sruvey <- list.files(model.dir.noradius_survey, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.fixedradius_survey <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_FixedRadius"
fixedradius.models_sruvey <- list.files(model.dir.fixedradius_survey, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.esw_survey <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_ESW"
esw.models_sruvey <- list.files(model.dir.esw_survey, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)


```


Initiate empty list
```{r}
# Species list
species_list <- unique(fixed.test_survey$Common_Name)

# Initialize lists with NA values
  # Observer-centered
true_labels_list_observer <- list()

  # Bird-centered
true_labels_list_bird <- list()


# No radius
bird_no_radius_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)
observer_no_radius_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)

# Fixed radius
bird_fixed_radius_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)
observer_fixed_radius_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)

# ESW
bird_esw_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)
observer_esw_preds_list <- setNames(rep(list(NA), length(species_list)), species_list)


```

Print prediction results
```{r}
# Modify the "Occur" column
noRadius.test_survey <- noRadius.test_survey %>%
  mutate(Occur = ifelse(Occur > 1, 1, Occur))

for (species_name in species_list) {
    
    # True labels for the species
    true_labels <- noRadius.test_survey[noRadius.test_survey$Common_Name == species_name, ]$Occur
    true_labels_list[[species_name]] <- true_labels
    
    #==== Bird-centered =====#
    # No radius
    no_radius_model_path <- grep(species_name, noradius.models_bird, value = TRUE)
    if (length(no_radius_model_path) > 0) {
        no_radius_model <- readRDS(no_radius_model_path)
        if (!is.null(no_radius_model)) {
            bird_no_radius_preds <- predict.gbm(object = no_radius_model,
                                                newdata = noRadius.test_bird[noRadius.test_bird$Common_Name == species_name, ],
                                                n.trees = no_radius_model$gbm.call$best.trees,
                                                type = "response")
            bird_no_radius_preds_list[[species_name]] <- bird_no_radius_preds
        }
    }
    
    # Fixed radius
    fixed_radius_model_path <- grep(species_name, fixedradius.models_bird, value = TRUE)
    if (length(fixed_radius_model_path) > 0) {
        fixed_radius_model <- readRDS(fixed_radius_model_path)
        if (!is.null(fixed_radius_model)) {
            bird_fixed_radius_preds <- predict.gbm(object = fixed_radius_model,
                                                   newdata = fixed.test_bird[fixed.test_bird$Common_Name == species_name, ],
                                                   n.trees = fixed_radius_model$gbm.call$best.trees,
                                                   type = "response")
            bird_fixed_radius_preds_list[[species_name]] <- bird_fixed_radius_preds
        }
    }
    
    # ESW
    esw_model_path <- grep(species_name, esw.models.models_bird, value = TRUE)
    if (length(esw_model_path) > 0) {
        esw_model <- readRDS(esw_model_path)
        if (!is.null(esw_model)) {
            bird_esw_preds <- predict.gbm(object = esw_model,
                                          newdata = esw.test_bird[esw.test_bird$Common_Name == species_name,],
                                          n.trees = esw_model$gbm.call$best.trees,
                                          type = "response")
            bird_esw_preds_list[[species_name]] <- bird_esw_preds
        }
    }
    
    # === Observer-centered ====#
    # No radius
    no_radius_observer_model_path <- grep(species_name, noradius.models_sruvey, value = TRUE)
    if (length(no_radius_observer_model_path) > 0) {
        no_radius_observer_model <- readRDS(no_radius_observer_model_path)
        if (!is.null(no_radius_observer_model)) {
            observer_no_radius_preds <- predict.gbm(object = no_radius_observer_model,
                                                    newdata = noRadius.test_survey[noRadius.test_survey$Common_Name == species_name, ],
                                                    n.trees = no_radius_observer_model$gbm.call$best.trees,
                                                    type = "response")
            observer_no_radius_preds_list[[species_name]] <- observer_no_radius_preds
        }
    }
    
    # Fixed radius
    fixed_radius_observer_model_path <- grep(species_name, fixedradius.models_sruvey, value = TRUE)
    if (length(fixed_radius_observer_model_path) > 0) {
        fixed_radius_observer_model <- readRDS(fixed_radius_observer_model_path)
        if (!is.null(fixed_radius_observer_model)) {
            observer_fixed_radius_preds <- predict.gbm(object = fixed_radius_observer_model,
                                                       newdata = fixed.test_survey[fixed.test_survey$Common_Name == species_name, ],
                                                       n.trees = fixed_radius_observer_model$gbm.call$best.trees,
                                                       type = "response")
            observer_fixed_radius_preds_list[[species_name]] <- observer_fixed_radius_preds
        }
    }
    
    # ESW
    esw_observer_model_path <- grep(species_name, esw.models_sruvey, value = TRUE)
    if (length(esw_observer_model_path) > 0) {
        esw_observer_model <- readRDS(esw_observer_model_path)
        if (!is.null(esw_observer_model)) {
            observer_esw_preds <- predict.gbm(object = esw_observer_model,
                                              newdata = esw.test_survey[esw.test_survey$Common_Name == species_name,],
                                              n.trees = esw_observer_model$gbm.call$best.trees,
                                              type = "response")
            observer_esw_preds_list[[species_name]] <- observer_esw_preds
        }
    }
}


```

[Observer-centered] Print prediction results
```{r}
# Modify the "Occur" column
noRadius.test_survey <- noRadius.test_survey %>%
  mutate(Occur = ifelse(Occur > 1, 1, Occur))

for (species_name in species_list) {
    
    # True labels for the species
    true_labels <- noRadius.test_survey[noRadius.test_survey$Common_Name == species_name, ]$Occur
    true_labels_list_observer[[species_name]] <- true_labels
    
    
    # === Observer-centered ====#
    # No radius
    no_radius_observer_model_path <- grep(species_name, noradius.models_sruvey, value = TRUE)
    if (length(no_radius_observer_model_path) > 0) {
        no_radius_observer_model <- readRDS(no_radius_observer_model_path)
        if (!is.null(no_radius_observer_model)) {
            observer_no_radius_preds <- predict.gbm(object = no_radius_observer_model,
                                                    newdata = noRadius.test_survey[noRadius.test_survey$Common_Name == species_name, ],
                                                    n.trees = no_radius_observer_model$gbm.call$best.trees,
                                                    type = "response")
            observer_no_radius_preds_list[[species_name]] <- observer_no_radius_preds
        }
    }
    
    # Fixed radius
    fixed_radius_observer_model_path <- grep(species_name, fixedradius.models_sruvey, value = TRUE)
    if (length(fixed_radius_observer_model_path) > 0) {
        fixed_radius_observer_model <- readRDS(fixed_radius_observer_model_path)
        if (!is.null(fixed_radius_observer_model)) {
            observer_fixed_radius_preds <- predict.gbm(object = fixed_radius_observer_model,
                                                       newdata = fixed.test_survey[fixed.test_survey$Common_Name == species_name, ],
                                                       n.trees = fixed_radius_observer_model$gbm.call$best.trees,
                                                       type = "response")
            observer_fixed_radius_preds_list[[species_name]] <- observer_fixed_radius_preds
        }
    }
    
    # ESW
    esw_observer_model_path <- grep(species_name, esw.models_sruvey, value = TRUE)
    if (length(esw_observer_model_path) > 0) {
        esw_observer_model <- readRDS(esw_observer_model_path)
        if (!is.null(esw_observer_model)) {
            observer_esw_preds <- predict.gbm(object = esw_observer_model,
                                              newdata = esw.test_survey[esw.test_survey$Common_Name == species_name,],
                                              n.trees = esw_observer_model$gbm.call$best.trees,
                                              type = "response")
            observer_esw_preds_list[[species_name]] <- observer_esw_preds
        }
    }
}


```

[Bird-location] Print prediction results
```{r}
# Modify the "Occur" column
noRadius.test_bird <- noRadius.test_bird %>%
  mutate(Occur = ifelse(Occur > 1, 1, Occur))

for (species_name in species_list) {
    
    # True labels for the species
    true_labels <- noRadius.test_bird[noRadius.test_bird$Common_Name == species_name, ]$Occur
    true_labels_list_bird[[species_name]] <- true_labels
    
    #==== Bird-centered =====#
    # No radius
    no_radius_model_path <- grep(species_name, noradius.models_bird, value = TRUE)
    if (length(no_radius_model_path) > 0) {
        no_radius_model <- readRDS(no_radius_model_path)
        if (!is.null(no_radius_model)) {
            bird_no_radius_preds <- predict.gbm(object = no_radius_model,
                                                newdata = noRadius.test_bird[noRadius.test_bird$Common_Name == species_name, ],
                                                n.trees = no_radius_model$gbm.call$best.trees,
                                                type = "response")
            bird_no_radius_preds_list[[species_name]] <- bird_no_radius_preds
        }
    }
    
    # Fixed radius
    fixed_radius_model_path <- grep(species_name, fixedradius.models_bird, value = TRUE)
    if (length(fixed_radius_model_path) > 0) {
        fixed_radius_model <- readRDS(fixed_radius_model_path)
        if (!is.null(fixed_radius_model)) {
            bird_fixed_radius_preds <- predict.gbm(object = fixed_radius_model,
                                                   newdata = fixed.test_bird[fixed.test_bird$Common_Name == species_name, ],
                                                   n.trees = fixed_radius_model$gbm.call$best.trees,
                                                   type = "response")
            bird_fixed_radius_preds_list[[species_name]] <- bird_fixed_radius_preds
        }
    }
    
    # ESW
    esw_model_path <- grep(species_name, esw.models.models_bird, value = TRUE)
    if (length(esw_model_path) > 0) {
        esw_model <- readRDS(esw_model_path)
        if (!is.null(esw_model)) {
            bird_esw_preds <- predict.gbm(object = esw_model,
                                          newdata = esw.test_bird[esw.test_bird$Common_Name == species_name,],
                                          n.trees = esw_model$gbm.call$best.trees,
                                          type = "response")
            bird_esw_preds_list[[species_name]] <- bird_esw_preds
        }
    }
    
   
        }

```


Check data consistency for each species
```{r}
# Observer-location
for (i in 1:length(true_labels_list_observer)) {
  cat("Species:", names(true_labels_list_observer)[i], "\n")
  cat("True Labels Length (observer):", length(true_labels_list_observer[[i]]), "\n")
  cat("Observer-location no radius predictions length:", length(observer_no_radius_preds_list[[i]]), "\n")
  cat("Observer-location fixed radius predictions length:", length(observer_fixed_radius_preds_list[[i]]), "\n")
  cat("Observer-location ESW predictions length:", length(observer_esw_preds_list[[i]]), "\n")
  cat("Unique True Labels:", unique(true_labels_list_observer[[i]]), "\n\n")
}

# Bird-location
for (i in 1:length(true_labels_list_bird)) {
  cat("Species:", names(true_labels_list_bird)[i], "\n")
  cat("True Labels Length (bird):", length(true_labels_list_bird[[i]]), "\n")
  cat("Bird-location no radius predictions length:", length(bird_no_radius_preds_list[[i]]), "\n")
  cat("Bird-location fixed radius predictions length:", length(bird_fixed_radius_preds_list[[i]]), "\n")
  cat("Bird-location ESW predictions length:", length(bird_esw_preds_list[[i]]), "\n")
  cat("Unique True Labels:", unique(true_labels_list_bird[[i]]), "\n\n")

}

```

DeLong test individual species
```{r}

roc1.obs <- roc(true_labels_list_observer[[6]], observer_no_radius_preds_list[[6]])
roc2.obs <- roc(true_labels_list_observer[[6]], observer_fixed_radius_preds_list[[6]])

roc.test(roc1.obs, roc2.obs, method = "delong")
```

Perform DeLong AUC tests
[Observer-location]
```{r}
perform_delong_tests_obs <- function(observer_preds_list_first, observer_preds_list_sec) {
  significant_count <- 0
  non_significant_count <- 0
  
  for (i in 1:length(true_labels_list_observer)) {
    true_labels <- true_labels_list_observer[[i]]
    
    if (length(unique(true_labels)) == 2) {
    
    roc_first <- tryCatch(roc(true_labels, observer_preds_list_first[[i]]), error = function(e) e)
    roc_sec <- tryCatch(roc(true_labels, observer_preds_list_sec[[i]]), error = function(e) e)
    
    if (inherits(roc_first, "error") || inherits(roc_sec, "error")) {
      cat("Skipping species due to ROC analysis error -", names(true_labels_list_observer)[i], "\n")
    } else {
    
    test_result <- roc.test(roc_first, roc_sec, method = "delong")
    
    if (test_result$p.value < 0.05) {
      significant_count <- significant_count + 1
    } else {
      non_significant_count <- non_significant_count + 1
      }
    }
    }else {
        cat("Skipping species", names(true_labels_list_observer)[i], "- not enough data for ROC analysis\n")
    }
  }
  return(list(significant_count = significant_count, non_significant_count = non_significant_count))
}



```

Evaluate and print results
-- Only test within predictor extraction strategy --
```{r}
# Observer-location
cat("Strategy 1: No Radius vs. Fixed Radius - Observer-location\n")
result_no_radius_obs <- perform_delong_tests_obs(observer_fixed_radius_preds_list, observer_no_radius_preds_list)


#cat("\nStrategy 2: Fixed Radius\n")
#result_fixed <- perform_delong_tests(bird_fixed_radius_preds_list, bird_fixed_radius_preds_list, observer_fixed_radius_preds_list)
#cat("\nStrategy 3: ESW Radius\n")
#result_esw <- perform_delong_tests(bird_esw_preds_list, bird_esw_preds_list, observer_esw_preds_list)
```



Script from my collaborator
```{r}
# Number of birds and records
n_birds <- 10
n_rec <- 100

# Define the true labels for the birds
true_labels_list <- replicate(n_birds, sample(c(0, 1), n_rec, replace = TRUE, prob = c(0.9, 0.1)), simplify = FALSE)
```



# Define predicted probabilities for birds and observers for each scenario
# Modify the probabilities to create a mix of significant and non-significant results

```{r}
# Replace below lists with Test data predictions
bird_no_radius_preds_list <- true_labels_list
observer_no_radius_preds_list <- replicate(n_birds, sample(c(0, 1), n_rec, replace = TRUE, prob = c(0.1, 0.9)), simplify = FALSE)

bird_fixed_preds_list <- true_labels_list
observer_fixed_preds_list <- replicate(n_birds, sample(c(0, 1), n_rec, replace = TRUE, prob = c(0.2, 0.8)), simplify = FALSE) 

bird_esw_preds_list <- true_labels_list
observer_esw_preds_list <- replicate(n_birds, sample(c(0, 1), n_rec, replace = TRUE, prob = c(1.0, 0.0)), simplify = FALSE) 


```

```{r}
# Function to perform DeLong test for each bird and print the results
perform_delong_tests <- function(true_labels_list, bird_preds_list, observer_preds_list) {
  significant_count <- 0
  non_significant_count <- 0
  
  for (i in 1:length(true_labels_list)) {
    roc_bird <- roc(true_labels_list[[i]], bird_preds_list[[i]])
    roc_observer <- roc(true_labels_list[[i]], observer_preds_list[[i]])
    
    test_result <- roc.test(roc_bird, roc_observer, method = "delong")
    
    if (test_result$p.value < 0.05) {
      significant_count <- significant_count + 1
      #cat("Bird", i, ": Significant (p-value =", test_result$p.value, ")\n")
    } else {
      non_significant_count <- non_significant_count + 1
      #cat("Bird", i, ": Not Significant (p-value =", test_result$p.value, ")\n")
    }
  }
  
  return(list(significant_count = significant_count, non_significant_count = non_significant_count))
}

```

```{r}
# Perform DeLong tests for each scenario
cat("Strategy 1: No Radius\n")
result_no_radius <- perform_delong_tests(true_labels_list, bird_no_radius_preds_list, observer_no_radius_preds_list)
cat("\nStrategy 2: Fixed Radius\n")
result_fixed <- perform_delong_tests(true_labels_list, bird_fixed_preds_list, observer_fixed_preds_list)
cat("\nStrategy 3: ESW Radius\n")
result_esw <- perform_delong_tests(true_labels_list, bird_esw_preds_list, observer_esw_preds_list)

# Calculate percentages
percent_significant_no_radius <- (result_no_radius$significant_count / n_birds) * 100
percent_non_significant_no_radius <- (result_no_radius$non_significant_count / n_birds) * 100

percent_significant_fixed <- (result_fixed$significant_count / n_birds) * 100
percent_non_significant_fixed <- (result_fixed$non_significant_count / n_birds) * 100

percent_significant_esw <- (result_esw$significant_count / n_birds) * 100
percent_non_significant_esw <- (result_esw$non_significant_count / n_birds) * 100

# Print summary results
cat("\nPercentage of Species with Significant AUC Difference:\n")
cat("No Radius: ", percent_significant_no_radius, "% significant, ", percent_non_significant_no_radius, "% not significant\n")
cat("Fixed Radius: ", percent_significant_fixed, "% significant, ", percent_non_significant_fixed, "% not significant\n")
cat("ESW Radius: ", percent_significant_esw, "% significant, ", percent_non_significant_esw, "% not significant\n")
```
