[Observer location BRT Model]

Goal: Evaluate species distribution models for all species of birds from Oregon trained BRT

Load packages
```{r}
library(pROC)
library(foreach)
library(doParallel)
library(reticulate)
library(rgee)
library(ggplot2)
library(mice)
library(gbm)
library(terra)
library(raster)
library(dismo)
library(gbm)
library(tidyverse)
library(readr)
```

Read ESW test dataset
# We will treat esw file differently since parsing issue has occurred
```{r}

esw.test <- readr::read_csv("/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TestData/SurveyCentered/SurveyCentered_ESW.csv", col_names = TRUE)
spec(esw.test)
problems(esw.test)

esw.test <- esw.test[-c(39007,39008),]
```


Load Test Data for observer-centered
```{r}
# Define the base directory
testdata_dir <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TestData/SurveyCentered/"

# Define file names
file_names <- c("SurveyCentered_FixedRadius.csv", "SurveyCentered_noRadius.csv")

# Load and assign datasets to variables
fixed.test <- read.csv(paste0(testdata_dir, file_names[1]))
noRadius.test <- read.csv(paste0(testdata_dir, file_names[2]))


```

Load the Trained models for observer-centered strategy
```{r}
individual_species_oregon <- split(fixed.test, fixed.test$Common_Name)

model.dir.noradius <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_NoRadius"
noradius.models <- list.files(model.dir.noradius, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.fixedradius <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_FixedRadius"
fixedradius.models <- list.files(model.dir.fixedradius, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

model.dir.esw <- "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/BRT_train/TrainedModel/SurveyCentered/SurveyCentered_ESW"
esw.models <- list.files(model.dir.esw, pattern = "\\.rds$", full.names = TRUE, recursive = TRUE)

# Create empty dataframe to store evaluation results
result <- data.frame(Species = character(),
                     Extraction_Approach = character(),
                     Metric = character(),
                     No_radius = numeric(),
                     Fixed_Radius = numeric(),
                     Effective_Strip_Width = numeric())

```

***===Model Evaluation (AUC, Kappa, True Skill Statistic (TSS))====***
***Assign NA to NULL BRT models for each species & predictor extraction method***
Load all models for a given species and calculate AUC values
```{r}
load_models_for_species <- function(species_name) {

    # Initialize variables to store AUC values
    auc_noRadius <- NA
    auc_fixed <- NA
    auc_esw <- NA

    # Initialize variables to store Kappa values
    kappa_noRadius <- NA
    kappa_fixed <- NA
    kappa_esw <- NA
    
    # Initialize variables to store TSS values
    tss_noRadius <- NA
    tss_fixed <- NA
    tss_esw <- NA
    
    #=== For no radius ===#
    noradius.perspecies <- grep(species_name, noradius.models, value = TRUE)
    
    if (length(noradius.perspecies) > 0) {
        noradius.perspecies.model <- readRDS(noradius.perspecies)
        
        if (!is.null(noradius.perspecies.model$gbm.call$best.trees)) {
            pred.noRadius <- predict.gbm(object = noradius.perspecies.model, 
                                         newdata = noRadius.test[noRadius.test$Common_Name == species_name, ],
                                         n.trees = noradius.perspecies.model$gbm.call$best.trees,
                                         type = "response")
                                         
            # Calculate AUC
            d.noRadius <- cbind(noRadius.test[noRadius.test$Common_Name == species_name, ]$Occur, pred.noRadius)
            pres.noRadius <- d.noRadius[d.noRadius[, 1] >= 1, 2]
            abs.noRadius <- d.noRadius[d.noRadius[, 1] == 0, 2]
            e.noRadius <- evaluate(p = pres.noRadius, a = abs.noRadius)
            auc_noRadius <- e.noRadius@auc
            
            # Calculate kappa
            kappa_noRadius <- max(e.noRadius@kappa)
            
            # Calculate TSS
            tss_noRadius <- max(e.noRadius@TPR + e.noRadius@TNR) - 1
        }
    }

    #=== For fixed radius ===#
    fixedradius.perspecies <- grep(species_name, fixedradius.models, value = TRUE)
    
    if (length(fixedradius.perspecies) > 0) {
        fixedradius.perspecies.model <- readRDS(fixedradius.perspecies)
        
        if (!is.null(fixedradius.perspecies.model$gbm.call$best.trees)) {
            pred.fixed <- predict.gbm(object = fixedradius.perspecies.model, 
                                      newdata = fixed.test[fixed.test$Common_Name == species_name, ],
                                      n.trees = fixedradius.perspecies.model$gbm.call$best.trees,
                                      type = "response")
                                      
            # Calculate AUC
            d.fixed <- cbind(fixed.test[fixed.test$Common_Name == species_name, ]$Occur, pred.fixed)
            pres.fixed <- d.fixed[d.fixed[, 1] >= 1, 2]
            abs.fixed <- d.fixed[d.fixed[, 1] == 0, 2]
            e.fixed <- evaluate(p = pres.fixed, a = abs.fixed)
            auc_fixed <- e.fixed@auc
            
             # Calculate kappa
            kappa_fixed <- max(e.fixed@kappa)
            
            # Calculate TSS
            tss_fixed <- max(e.fixed@TPR + e.fixed@TNR) - 1
        }
    }

    #=== For ESW ===#
    esw.perspecies <- grep(species_name, esw.models, value = TRUE)
    
    if (length(esw.perspecies) > 0) {
        esw.perspecies.model <- readRDS(esw.perspecies)
        
        if (!is.null(esw.perspecies.model$gbm.call$best.trees)) {
            pred.ESW <- predict.gbm(object = esw.perspecies.model, 
                                    newdata = esw.test[esw.test$Common_Name == species_name, ],
                                    n.trees = esw.perspecies.model$gbm.call$best.trees,
                                    type = "response")
                                    
            # Calculate AUC
            d.esw <- cbind(esw.test[esw.test$Common_Name == species_name, ]$Occur, pred.ESW)
            pres.ESW <- d.esw[d.esw[, 1] >= 1, 2]
            abs.ESW <- d.esw[d.esw[, 1] == 0, 2]
            e.esw <- evaluate(p = pres.ESW, a = abs.ESW)
            auc_esw <- e.esw@auc
            
           # Calculate kappa
            kappa_esw <- max(e.esw@kappa)
            
            # Calculate TSS
            tss_esw <- max(e.esw@TPR + e.esw@TNR) - 1
        }
    }

    # Store results in a dataframe
     # AUC
    per_bird.result.AUC <- data.frame(Species = species_name,
                                  Extraction_Approach = "Observer-Centered Approach",
                                  Metric = "AUC",
                                  No_radius = auc_noRadius,
                                  Fixed_Radius = auc_fixed,
                                  Effective_Strip_Width = auc_esw)
    # Kappa
    per_bird.result.kappa <- data.frame(Species = species_name,
                                  Extraction_Approach = "Observer-Centered Approach",
                                  Metric = "Kappa",
                                  No_radius = kappa_noRadius,
                                  Fixed_Radius = kappa_fixed,
                                  Effective_Strip_Width = kappa_esw)
    
    # TSS
    per_bird.result.tss <- data.frame(Species = species_name,
                                  Extraction_Approach = "Observer-Centered Approach",
                                  Metric = "TSS",
                                  No_radius = tss_noRadius,
                                  Fixed_Radius = tss_fixed,
                                  Effective_Strip_Width = tss_esw)
    
    # Concatenate per bird results in one data frame
    per_bird.result <- rbind(per_bird.result.AUC, per_bird.result.kappa, per_bird.result.tss)
    
    return(per_bird.result)
}


# Loop over each bird species name and load models
for (bird in names(individual_species_oregon)) {
  print(bird)
  per_bird.result <- load_models_for_species(bird)
  result <- rbind(result, per_bird.result)
}

```

Export as CSV
```{r}
write.table(result, file = "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/Evaluation/evalMetric-ObserverCentered.csv", sep = ",", row.names = FALSE, fileEncoding = "UTF-8")

```

Print & plot individual results
```{r}
print(result)
tapply(result[4:6], result$Metric, summary)


# Plot AUC box plot
AUC <- result[result$Metric == "AUC",]
png(filename = "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/Evaluation/Figures/AUC-ObserverCentered.png", width = 2500, height = 2200, res = 300)
boxplot(AUC[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "AUC",
        main = "Observer-centered Landsat bands predictor mapping strategy evaluation")
dev.off()

# Plot Kappa
kappa <- result[result$Metric == "Kappa",]
png(filename = "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/Evaluation/Figures/Kappa-ObserverCentered.png", width = 2500, height = 2200, res = 300)
boxplot(kappa[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "Kappa",
        main = "Observer-centered Landsat bands predictor mapping strategy evaluation")
dev.off()

# Plot TSS
tss <- result[result$Metric == "TSS",]
png(filename = "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/Evaluation/Figures/TSS-ObserverCentered.png", width = 2500, height = 2200, res = 300)
boxplot(tss[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "TSS",
        main = "Observer-centered Landsat bands predictor mapping strategy evaluation")
dev.off()


```


Combine evaluation metrics together
```{r}
# Plot AUC box plot
AUC <- result[result$Metric == "AUC",]
kappa <- result[result$Metric == "Kappa",]
tss <- result[result$Metric == "TSS",]

png(filename = "/Volumes/T7/eBird_Oregon2020/distance/Distance-Sampling/Evaluation/Figures/evaMetic-ObserverCentered.png", width = 4500, height = 4200, res = 300)
par(mfrow = c(2,2))
 
  # AUC
boxplot(AUC[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "AUC",
        main = "Observer-centered AUC evaluation")

  # Kappa
boxplot(kappa[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "Kappa",
        main = "Observer-centered Cohen's Kappa evaluation")


  # TSS
boxplot(tss[,c(4:6)],
        xlab = "Mapping strategy",
        ylab = "TSS",
        main = "Observer-centered True Skill Statistic evaluation")
dev.off()

```
